---
title: "Data Visualization"
author: "Adejumo Ridwan Suleiman"
format: revealjs
---


## What is Data Visualization?

- Graphical representation of data using charts and plots.
- Easily communicate information making it easy to understand.

## Types of Data Visualization

- Histogram

- Frequency Polygon

- Stem and Leaf Plots

- Box Plots

- Bar chart

- Pie chart

# Histogram

## Histogram

- Used with discrete or continuous data
- Shows the underlying distribution of a data
- The data is split into intervals called *bins*
- Each bin contains the number of occurrences of scores in the data set contained within that bin.

## Histogram

```{r echo=FALSE}
# Load ggplot2 package
library(ggplot2)

# Load the mtcars dataset
data("mtcars")

# Create histogram of the mpg variable
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Miles Per Gallon (mpg) in mtcars Dataset",
       x = "Miles Per Gallon (mpg)",
       y = "Frequency") +
  theme_minimal()

```


# Bar Chart

## Bar Chart

- Used with categorical data
- Bars are either plotted horizontally or vertically
- Compares various categories

## Bar Chart

```{r echo=FALSE}
# Load the ggplot2 package
library(ggplot2)

# Create a bar chart colored by diamond cut
ggplot(data = diamonds, aes(x = cut, fill = cut)) + 
  geom_bar(color = "black") + 
  labs(title = "Number of Diamonds by Cut",
       x = "Cut",
       y = "Count of Diamonds") + 
  theme_minimal()

```

## Difference between histogram and bar chart

| Feature | Histogram | Bar Chart |
| --- | --- | --- |
|Data type | Used with numerical data | Used with categorical data |
| Purpose | Represents the distribution of numerical data | Compares different categories or groups |
| Bars | Touch each other (no gaps) | Separated by spaces |

# Box Plot

## Box plot

- Also known as box and whisker plot
- Used to measure the measures of central tendency in a data
- Used with numerical data
- Divides a data into 4 parts

## Box Plot

```{r}
# Load the ggplot2 package
library(ggplot2)

# Load the iris dataset
data(iris)

# Create a boxplot with color by species
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of Sepal Length by Species",
       x = "Species",
       y = "Sepal Length")


```


## Box Plot

![Different parts of a boxplot. Source: [BYJU](https://byjus.com/maths/box-plot/)](images/boxplot.jpg)


# Pie Chart

## Pie Chart


# Scatter plot


# Line plot


# Frequency Polygon


# Stem and Leaf Plot
