---
title: "Data Visualization"
author: "Adejumo Ridwan Suleiman"
format: revealjs
---


## What is Data Visualization?

- Graphical representation of data using charts and plots.
- Easily communicate information making it easy to understand.

## Types of Data Visualization

- Histogram

- Bar Chart

- Box plots

- Pie Chart

- Scatter plot

- Line Graph

- Frequency Polygon

# Histogram

## Histogram

- Used with discrete or continuous data
- Shows the underlying distribution of a data
- The data is split into intervals called *bins*
- Each bin contains the number of occurrences of scores in the data set contained within that bin.

## Histogram

```{r echo=FALSE}
# Load ggplot2 package
library(ggplot2)

# Load the mtcars dataset
data("mtcars")

# Create histogram of the mpg variable
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Miles Per Gallon (mpg) in mtcars Dataset",
       x = "Miles Per Gallon (mpg)",
       y = "Frequency") +
  theme_minimal()

```


# Bar Chart

## Bar Chart

- Used with categorical data
- Bars are either plotted horizontally or vertically
- Compares various categories

## Bar Chart

```{r echo=FALSE}

# Create a bar chart colored by diamond cut
ggplot(data = diamonds, aes(x = cut, fill = cut)) + 
  geom_bar(color = "black") + 
  labs(title = "Number of Diamonds by Cut",
       x = "Cut",
       y = "Count of Diamonds") + 
  theme_minimal()

```

## Difference between histogram and bar chart

| Feature | Histogram | Bar Chart |
| --- | --- | --- |
|Data type | Used with numerical data | Used with categorical data |
| Purpose | Represents the distribution of numerical data | Compares different categories or groups |
| Bars | Touch each other (no gaps) | Separated by spaces |

# Box Plot

## Box plot

- Also known as box and whisker plot
- Used to measure the measures of central tendency in a data
- Used with numerical data
- Divides a data into 4 parts

## Box Plot

```{r echo=FALSE}


# Load the iris dataset
data(iris)

# Create a boxplot with color by species
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
  geom_boxplot() +
  labs(title = "Boxplot of Sepal Length by Species",
       x = "Species",
       y = "Sepal Length") + 
  theme_minimal()


```


## Box Plot

![Different parts of a boxplot. Source: [BYJU](https://byjus.com/maths/box-plot/)](images/boxplot.jpg)


# Pie Chart

## Pie Chart

- Circular graph divided into slices
- Used mostly with categorical data
- Each slice represents the frequency of a particular category in percentage
- The entire pie represents 100%

## Pie Chart

```{r  echo=FALSE}
library(dplyr)

# Use the diamonds dataset (comes with ggplot2)
data <- diamonds

# Create a data frame with cut counts and percentages
cut_counts <- data %>%
  count(cut) %>%
  mutate(percentage = n / sum(n))

# Create the pie chart
ggplot(cut_counts, aes(x = "", y = percentage, fill = cut)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Distribution of Diamonds by Cut",
       fill = "Cut") +
  theme_void() +
  geom_text(aes(label = paste0(round(percentage * 100, 1), "%")), 
            position = position_stack(vjust = 0.5))
```

# Scatter plot

## Scatter Plot

- Can be used to compare two continuous variables
- Used in also determining the relationship between two variables

## Scatter plot

```{r echo=FALSE}
# Create the scatterplot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Car Weight vs. Fuel Efficiency",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()
```


# Line Graph

## Line Graph

- Represents change in a variable with respect to time.
- For example, change in the population of a city in the past two decades.
- Used to represent quantitative data by plotting each variable value with respect to the time of occurence.


## Line Graph

```{r echo=FALSE}
# Load required library
library(ggplot2)

# Create the plot
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = "blue") +
  labs(title = "US Unemployment Over Time",
       x = "Date",
       y = "Number of Unemployed (Thousands)") +
  theme_minimal()
```


# Frequency Polygon

## Frequency Polygon

- Similar to a histogram
- Used to visualize the underlying distribution of a continuous variable


## Frequency Polygon

```{r echo=FALSE}
# Load the ggplot2 library
library(ggplot2)

# Create the frequency polygon
ggplot(diamonds, aes(x = price)) +
  geom_freqpoly(binwidth = 500, color = "blue") +
  labs(title = "Frequency Polygon of Diamond Prices",
       x = "Price (USD)",
       y = "Frequency") +
  theme_minimal()
```

## Frequency polygon from histogram

```{r echo=FALSE}
# Create the histogram with frequency polygon
ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 500, fill = "lightblue", color = "black") +
  geom_freqpoly(binwidth = 500, color = "blue") +
  labs(title = "Histogram and Frequency Polygon of Diamond Prices",
       x = "Price (USD)",
       y = "Frequency") +
  theme_minimal()
```

