---
title: "Functions and Packages"
author: "Adejumo Ridwan Suleiman"
output: html_document
---

### Functions

A function is a ready-made set of instructions - code - that tells R to do something. There are thousands of functions in R. Typically, you insert a variable into the function, and it returns an answer. The code for doing this follows the pattern `function_name(variable_name)`. 

```{r eval=FALSE}
# Compute the mean age of bookstore customers
age <- c(28, 48, 47, 71, 22, 80, 48, 30, 31)
mean(age)
```

Note that the code follows the pattern `function_name(variable_name)`: the function's name is `mean` and the variable's name is `age`.


```{r eval=FALSE}
# Compute the correlation between the variables age and purchase
age <- c(28, 48, 47, 71, 22, 80, 48, 30, 31)
purchase <- c(20, 59, 2, 12, 22, 160, 34, 34, 29)
cor(age, purchase)
```



```{r eval=FALSE}
cor(age, purchase, method = "spearman")
```


So, how can we know what arguments to pass to a function? Luckily, we don't have to memorise all possible arguments for all functions. Instead, we can look at the _documentation_, i.e. help file, for a function that we are interested in. This is done by typing `?function_name` in the Console panel, or doing a web search for `R function_name`. To view the documentation for the `cor` function, type:

```{r eval=FALSE}
?cor
```


#### Exercise
Using the data you created in Exercise, do the following:
1. Compute the mean height of the people.
2. Compute the correlation between height and weight.

Do the following:
1. Read the documentation for the function `length`. What does it do? Apply it to your `height` vector.
2. Read the documentation for the function `sort`. What does it do? What does the argument `decreasing` (the values of which can be either `FALSE` or `TRUE`) do? Apply the function to your `weight` vector.
