---
title: "DataFrames"
author: "Adejumo Ridwan Suleiman"
output: html_document
---

## Vectors in data frames
In the next few sections, we will explore the `airquality` dataset. It contains daily air quality measurements from New York during a period of five months:

* `Ozone`: mean ozone concentration (ppb),
* `Solar.R`: solar radiation (Langley),
* `Wind`: average wind speed (mph),
* `Temp`: maximum daily temperature in degrees Fahrenheit,
* `Month`: numeric month (May=5, June=6, and so on),
* `Day`: numeric day of the month (1-31).

There are lots of things that would be interesting to look at in this dataset. What was the mean temperature during the period? Which day was the hottest? Which was the windiest? What days were the temperature more than 90 degrees Fahrenheit? To answer these questions, we need to be able to access the vectors inside the data frame. We also need to be able to quickly and automatically screen the data in order to find interesting observations (e.g. the hottest day)

### Accessing vectors and elements {#accessingelements}
In Section, we learned how to compute the mean of a vector. We also learned that to compute the mean of a vector _that is stored inside a data frame_ we could use a dollar sign: `data_frame_name$vector_name`. Here is an example with the `airquality` data:

```{r eval=FALSE}

```

If we want to grab a particular element from a vector, we must use its _index_ within square brackets: `[index]`. The first element in the vector has index 1, the second has index 2, the third index 3, and so on. To access the fifth element in the `Temp` vector in the `airquality` data frame, we can use:

```{r eval=FALSE}
]
```

The square brackets can also be applied directly to the data frame. The syntax for this follows that used for matrices in mathematics: `airquality[i, j]` means the element at the i:th row and j:th column of `airquality`. We can also leave out either `i` or `j` to extract an entire row or column from the data frame. Here are some examples:

```{r eval=FALSE}

```



### Use your dollars
The `$` operator can be used not just to extract data from a data frame, but also to manipulate it. Let's return to our `bookstore` data frame, and see how we can make changes to it using the dollar sign.

```{r eval=FALSE}

```

Perhaps there was a data entry error - the second customer was actually 18 years old and not 48. We can assign a new value to that element by referring to it in either of two ways:

```{r eval=FALSE}

```

We could also change an entire column if we like. For instance, if we wish to change the `age` vector to months instead of years, we could use
```{r eval=FALSE}

```

What if we want to add another variable to the data, for instance the length of the customers' visits in minutes? There are several ways to accomplish this, one of which involves the dollar sign:

```{r eval=FALSE}

```

As you see, the new data has now been added to a new column in the data frame.




### Using conditions 
A few paragraphs ago, we were asking which was the hottest day in the `airquality` data. Let's find out! We already know how to find the maximum value in the `Temp` vector:

```{r eval=FALSE}

```

But can we find out which day this corresponds to? We could of course manually go through all 153 days e.g. by using `View(airquality)`, but that seems tiresome and wouldn't even be possible in the first place if we'd had more observations. A better option is therefore to use the function `which.max`:

```{r eval=FALSE}

```

`which.max` returns the index of the observation with the maximum value. If there is more than one observation attaining this value, it only returns the first of these.

We've just used `which.max` to find out that day `120` was the hottest during the period. If we want to have a look at the entire row for that day, we can use

```{r eval=FALSE}

```

Alternatively, we could place the call to `which.max` inside the brackets. Because `which.max(airquality$Temp)` returns the number `120`, this yields the same result as the previous line:

```{r eval=FALSE}

```

Were we looking for the day with the lowest temperature, we'd use `which.min` analogously. In fact, we could use any function or computation that returns an index in the same way, placing it inside the brackets to get the corresponding rows or columns. This is extremely useful is we want to extract observations with certain properties, for instance all days where the temperature was above 90 degrees. We do this using _conditions_, i.e. by giving statements that we wish to be fulfilled.

As a first example of a condition, we use the following, which checks if the temperature exceeds 90 degrees:

```{r eval=FALSE}

```

For each element in `airquality$Temp` this returns either `TRUE` (if the condition is fulfilled, i.e. when the temperature is greater than 90) or `FALSE` (if the conditions isn't fulfilled, i.e. when the temperature is 90 or lower). If we place the condition inside brackets following the name of the data frame, we will extract only the rows corresponding to those elements which were marked with `TRUE`:

```{r eval=FALSE}

```

If you prefer, you can also store the `TRUE` or `FALSE` values in a new variable:

```{r eval=FALSE}

```

There are several logical operators and functions which are useful when stating conditions in R:

```{r eval=FALSE}
a <- 3
b <- 8
a == b     # Check if a equals b
a > b      # Check if a is greater than b
a < b      # Check if a is less than b
a >= b     # Check if a is equal to or greater than b
a <= b     # Check if a is equal to or less than b
a != b     # Check if a is not equal to b
is.na(a)   # Check if a is NA
a %in% c(1, 4, 9) # Check if a equals at least one of 1, 4, 9
```  

When checking a conditions for all elements in a vector, we can use `which` to get the indices of the elements that fulfill the condition:

```{r eval=FALSE}
which(airquality$Temp > 90)
```  

If we want to know if all elements in a vector fulfill the condition, we can use `all`

```{r eval=FALSE}
all(airquality$Temp > 90)
```  

In this case, it returns `FALSE`, meaning that not all days had a temperature above 90 (phew!). Similarly, if we wish to know whether _at least one_ day had a temperature above 90, we can use `any`:

```{r eval=FALSE}
any(airquality$Temp > 90)
```

To find how many elements that fulfill a condition, we can use `sum`:

```{r eval=FALSE}
sum(airquality$Temp > 90)
```

Why does this work? Remember that `sum` computes the sum of the elements in a vector, and that when `logical` values are used in computations, they are treated as `0` (`FALSE`) or `1` (`TRUE`). Because the condition returns a vector of `logical` values, the sum of them becomes the number of 1's - the number of `TRUE` values  - i.e. the number of elements that fulfill the condition.

To find the proportion of elements that fulfill a condition, we can count how many elements fulfill it and then divide by how many elements are in the vector. This is exactly what happens if we use `mean`:

```{r eval=FALSE}
mean(airquality$Temp > 90)
```

Finally, we can combine conditions by using the logical operators `&` (AND), `|` (OR). Here are some examples:

```{r eval=FALSE}
a <- 3
b <- 8
# Is a less than b and greater than 1?
a < b & a > 1
# Is a less than b and equal to 4?
a < b & a == 4
# Is a less than b and/or equal to 4?
a < b | a == 4
# Is a equal to 4 and/or equal to 5?
a == 4 | a == 5
```



